---
title: "Dataverse Metrics Report"
date: "Generated `r Sys.time()`"
output:
  pdf_document: 
    includes:
      in_header: header.tex
  html_document: default
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE) 
opts <- options(knitr.kable.NA = "")
library(httr)
library(readr)
library(tidyr)
library(dplyr)
library(ggplot2)
library(ggfortify)
source("R/deposits.R")
source("R/mdc.R")

### Configurable variables

# Custom start and end dates: use YYYY-MM-DD format
# These are specified from the workflow
# If these aren't provided, we'll: default to today's date as end date and to the beginning of the year as the start date
period_begin <- as.Date(Sys.getenv("period_begin"))
period_end <- as.Date(Sys.getenv("period_end"))

# Calculate dates: We use YTD if none are given
if (is.na(period_begin)) {
  period_begin <- as.Date(paste(as.integer(format(Sys.Date(), "%Y")),
                              "01-01", sep = "-"))}
if (is.na(period_end)){
  period_end <- Sys.Date()
}

# Set tokens and file location settings
dataverse_host <- Sys.getenv("DATAVERSE_SERVER")
dataverse_key <- Sys.getenv("DATAVERSE_TOKEN")

dataverse <- paste0('http://', dataverse_host, "/api/info/metrics")
doi_link <- paste0('http://', dataverse_host, "/dataset.xhtml?persistentId=")

# Dataverse data. Cached locally as rds files
# This way, we don't need to re-scrape for each format (HTML/PDF)
if (file.exists("deposits.rds")) {
  deposits <- readRDS("deposits.rds")
  projects <- readRDS("projects.rds")
  if (file.exists("downloads.rds")) downloads <- readRDS("downloads.rds")
  if (file.exists("mdc.rds")) mdc <- readRDS("mdc.rds")
} else { # actually scrape
  deposits <- deposits_get(dataverse_host)
  saveRDS(deposits, "deposits.rds")

  projects <- content(GET(paste0(dataverse, "/uniquedownloads/monthly")))
  projects <- bind_rows(projects$data)
  saveRDS(projects, "projects.rds")

  if (mdc_available(dataverse_host)) {
    mdc <- mdc_get(dataverse_host)
    saveRDS(mdc, "mdc.rds")
  } else {
    downloads <- content(GET(paste0(dataverse, "/downloads/monthly")))
    downloads <- bind_rows(downloads$data)
    downloads <- projects %>% group_by(date) %>% summarize(unique=sum(count)) %>%
      full_join(downloads)
    saveRDS(downloads, "downloads.rds")
  }
}
# Always Fix inconsistent subject types 
deposits$subjects <- lapply(deposits$subjects, function(x) { 
  if (is.null(x) || is.logical(x)) return(NA_character_) 
      as.character(x) 
 })


```

## Parameters

- Dataverse instance: `r dataverse_host`
- Reporting period: `r period_begin` to `r period_end`.

```{r total_projects}
dataverse_total <- nrow(deposits)

# all projects created within specified time frame
deposits_new <- deposits %>% filter(as.Date(createdAt) < period_end & as.Date(createdAt) > period_begin) %>%
  summarize(n=n()) %>% .$n

# all *published* projects within the specified timeframe
published_new <- deposits %>% filter(as.Date(published_at) < period_end & as.Date(published_at) > period_begin) %>%
  summarize(n=n()) %>% .$n
```

- There are currently `r dataverse_total` projects. 
- `r deposits_new` new deposits were initiated between `r period_begin` and `r period_end`.
- `r published_new` projects were published between `r period_begin` and `r period_end`.

```{r subject}
# all projects by status
deposits %>% group_by(versionState) %>% summarize(n=n()) %>%
  knitr::kable(caption="Total deposits by publication status", col.names=c("Status", "Deposits"))


# helper to safely unnest and coerce
as_char_subjects <- function(df) {
  df %>%
    unnest_longer(subjects, keep_empty = TRUE) %>% 
    mutate(subjects = as.character(subjects)) %>%
    tidyr::replace_na(list(subjects = "Unknown"))
}

# 1) Total published by subject (all time)
published_all <- deposits %>%
  filter(versionState == "Published") %>%
  select(subjects) %>%
  as_char_subjects() %>%
  group_by(subjects) %>%
  summarize(published = n(), .groups = "drop")

# 2) Published in period by subject
published_period <- deposits %>%
  filter(versionState == "Published",
         as.Date(published_at) < period_end,
         as.Date(published_at) > period_begin) %>%
  select(subjects) %>%
  as_char_subjects() %>%
  group_by(subjects) %>%
  summarize(period = n(), .groups = "drop")

# 3) Unpublished totals by subject (if any)
if (any(deposits$versionState == "Unpublished")) {
  unpublished_all <- deposits %>%
    filter(versionState == "Unpublished") %>%
    select(subjects) %>%
    as_char_subjects() %>%
    group_by(subjects) %>%
    summarize(unpublished = n(), .groups = "drop")
} else {
  unpublished_all <- tibble(subjects = character(), unpublished = integer())
}

# 4) Unpublished in period by subject (if any)
if (any(subset(deposits,
               as.Date(createdAt) < period_end & as.Date(createdAt) > period_begin)$versionState == "Unpublished")) {
  unpublished_period <- deposits %>%
    filter(versionState == "Unpublished",
           as.Date(createdAt) < period_end,
           as.Date(createdAt) > period_begin) %>%
    select(subjects) %>%
    as_char_subjects() %>%
    group_by(subjects) %>%
    summarize(unpublished_period = n(), .groups = "drop")
} else {
  unpublished_period <- tibble(subjects = character(), unpublished_period = integer())
}

# Combine (all joins now match on a character 'subjects')
deposits_subj <- published_all %>%
  full_join(published_period, by = "subjects") %>%
  full_join(unpublished_all, by = "subjects") %>%
  full_join(unpublished_period, by = "subjects") %>%
  arrange(subjects)

# Columns depending on token presence
if (identical(dataverse_key, "") || is.na(dataverse_key)) {
  include_columns <- c("subjects", "period", "published")
  column_names <- c("Subject",
                    paste("Published between", period_begin, "and", period_end),
                    "Total published")
} else {
  include_columns <- c("subjects", "unpublished_period", "unpublished", "period", "published")
  column_names <- c("Subject",
                    paste("Unpublished between", period_begin, "and", period_end),
                    "Total draft",
                    paste("Published between", period_begin, "and", period_end),
                    "Total published")
}

deposits_subj[, include_columns] %>%
  tidyr::replace_na(list(
    unpublished_period = 0L,
    unpublished        = 0L,
    period             = 0L,
    published          = 0L
  )) %>%
  knitr::kable(caption = "Deposits by subject",
               col.names = column_names)


```

```{r top}
projects %>% group_by(pid) %>% summarize(sum(count)) %>% 
  arrange(desc(`sum(count)`)) %>% slice(1:5) %>%
  mutate(pid = paste0("[", deposits$name[deposits$global_id %in% pid], "](", 
                      doi_link, pid, ")")) %>%
  knitr::kable(caption="Top 5 all time downloaded projects",
             row.names = FALSE, col.names = c("Project", "Downloads"))

projects$date <- as.Date(paste0(projects$date, "-01"))
projects %>% group_by(pid) %>% filter(date < period_end &
                      date > period_begin) %>%
  summarize(sum(count)) %>% arrange(desc(`sum(count)`)) %>% slice(1:5) %>%
  mutate(pid = paste0("[", deposits$name[deposits$global_id %in% pid], "](",
                      doi_link, pid, ")")) %>%
  knitr::kable(caption=paste("Top 5 downloaded projects between ", period_begin,
                             "and", period_end),
             row.names = FALSE, col.names = c("Project", "Downloads"))
```

## Aggregate data

Dataverse reports statistics as "total" and "unique". 

- Unique" views/downloads correspond to the number of unique sessions in which a data project has been viewed/downloaded. Roughly translates to: "X people have viewed/downloaded this project or its files."
- Total views/downloads correspond to the number of times a project has been viewed/downloaded, including, e.g., multiple counts for multiple views/downloads in the same session or by the same user.

```{r mdc}
# not all dataverse instances use MDC data (QDR does). In case they don't, we use the regular downloads stats
if (exists("mdc")) {
  stats_ts <- ts(mdc[, -1], frequency=12, start=strsplit(mdc$date[1], split="-")[[1]])
} else {
  stats_ts <- ts(downloads[, -1], frequency=12, 
                start=strsplit(downloads$date[1], split="-")[[1]])
}

autoplot(stats_ts, facets = FALSE) +  labs(title="Statistics", color="Statistic") +
    scale_y_continuous(labels=scales::comma)

```

